On the effects of partying on next-day sports performance

### 2017-07-03 writing/draft.tex
Updated based on MSA analysis.

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the county in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the total number
of musical groups, sound recording studios, and musical publishers in county of the 
last game location, else emit zero.  We also re-scale this feature to lie within the
interval $[0,1]$ such that the magnitude of the coefficient is comparable with our
discrete indicator of nightlife.
This captures how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

\paragraph{How an unknown spread variable may (not) affect our inference}
For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically mis-values
teams which are more likely to be encountered after playing a city with active nightlife,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reasonably reliable. 
So, if we do find an effect in our variable of interest,
we can be reasonably confident in our conclusion that there is a causal effect and not some other systematic bias.

\subsection{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the county-quarter level. We obtain this data for each county corresponding
to an NBA or MLB team from years 2010-2016. 
As a proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular county-year: we simply calculate the total number of establishments across all three music categories listed above. We caveat that since the Toronto Raptors are located in Canada, we don't have BLS data for this team.\footnote{We don't expect this to be a party city.}

@@

\subsection{Results}
\subsubsection{NBA: using points-spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, recovery time, and home-team effect that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.
In laymens terms, teams under perform book-maker's expectations after playing in cities 
with active nightlife.

\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Meet the Spread} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]   Party discrete & $-$0.513$^{***}$ &  \\    & (0.145) &  \\    & & \\   Party continuous &  & $-$0.199$^{**}$ \\    &  & (0.100) \\    & & \\   Lag changes in posession & 0.0001 & 0.00004 \\    & (0.002) & (0.002) \\    & & \\   Logged travel distance & $-$0.017 & $-$0.014 \\    & (0.026) & (0.026) \\    & & \\   East-west travel direction & $-$0.146 & $-$0.112 \\    & (0.231) & (0.231) \\    & & \\   Number hours rest time & 0.001 & 0.001 \\    & (0.001) & (0.001) \\    & & \\   Time of game during day & 0.005 & 0.004 \\    & (0.013) & (0.013) \\    & & \\   Home team effect & 0.001 & 0.003 \\    & (0.043) & (0.043) \\    & & \\   Logged travel distance * east-west & 0.022 & 0.017 \\    & (0.036) & (0.036) \\    & & \\   Constant & $-$0.048 & $-$0.034 \\    & (0.378) & (0.378) \\    & & \\  \hline \\[-1.8ex]  Observations & 9,823 & 9,823 \\  Log Likelihood & $-$6,800.675 & $-$6,805.106 \\  Akaike Inf. Crit. & 13,619.350 & 13,628.210 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

\paragraph{Differences between NBA and MLB as it pertains to our analysis}
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. Interviews with athletes suggest that MLB players are more likely to
restrict their partying to weekends.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up is repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Probability of Winning} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]   Continuous measure of nightlife & $-$0.142$^{*}$ &  \\    & (0.083) &  \\    & & \\   Nightlife (no weekend interaction) &  & 0.004 \\    &  & (0.107) \\    & & \\   Bookmaker's odds & 2.600$^{***}$ & 2.604$^{***}$ \\    & (0.162) & (0.163) \\    & & \\   Home-team effect & 0.024 & 0.052 \\    & (0.035) & (0.072) \\    & & \\   Number of rest days & 0.015 & 0.015 \\    & (0.019) & (0.020) \\    & & \\   Logged travel distance & $-$0.001 & $-$0.001 \\    & (0.005) & (0.005) \\    & & \\   Weekend & 0.042 & 0.0004 \\    & (0.037) & (0.030) \\    & & \\   Constant & $-$1.329$^{***}$ & $-$1.348$^{***}$ \\    & (0.080) & (0.102) \\    & & \\  \hline \\[-1.8ex]  Observations & 25,215 & 24,781 \\  Log Likelihood & $-$17,274.540 & $-$16,977.790 \\  Akaike Inf. Crit. & 34,563.090 & 33,969.580 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

Indeed, we see that if players have enough recovery time after visiting cities such as Los
Angeles or New York, we cannot reject the null hypothesis that player performance is unaffected
by visitation to cities with active nightlife. The same is true in the MLB.

\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   party placebo & 0.072 \\    & (0.082) \\    & \\   lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   east-west travel direction & $-$0.100 \\    & (0.232) \\    & \\   number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   time of game during day & 0.001 \\    & (0.013) \\    & \\   home team effect & $-$0.009 \\    & (0.043) \\    & \\   logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.070 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.701 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Probability of Winning (MLB) \\  \hline \\[-1.8ex]   Party placebo & 0.004 \\    & (0.107) \\    & \\   Bookmaker's odds & 2.604$^{***}$ \\    & (0.163) \\    & \\   Home-team effect & 0.052 \\    & (0.072) \\    & \\   Number of rest days & 0.015 \\    & (0.020) \\    & \\   Logged travel distance & $-$0.001 \\    & (0.005) \\    & \\   Weekend & 0.0004 \\    & (0.030) \\    & \\   Constant & $-$1.348$^{***}$ \\    & (0.102) \\    & \\  \hline \\[-1.8ex]  Observations & 24,781 \\  Log Likelihood & $-$16,977.790 \\  Akaike Inf. Crit. & 33,969.580 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\subsection{Possible confounders or potential issues}

@@

\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Demeanead Rebounds Per Minute \\  \hline \\[-1.8ex]   Continuous party measure & $-$0.001$^{*}$ \\    & (0.0005) \\    & \\   Lag change in possessions & 0.00003 \\    & (0.0001) \\    & \\   Logged travel distance & 0.001 \\    & (0.001) \\    & \\   Number hours since last game & 0.00003 \\    & (0.00002) \\    & \\   Constant & $-$0.008$^{*}$ \\    & (0.004) \\    & \\  \hline \\[-1.8ex]  Observations & 43,547 \\  R$^{2}$ & 0.0003 \\  Adjusted R$^{2}$ & 0.0002 \\  Residual Std. Error & 0.104 (df = 43542) \\  F Statistic & 2.981$^{**}$ (df = 4; 43542) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

\hspace{-100pt}
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Points Admitted by Team & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.828$^{***}$ &  &  \\    & (0.699) &  &  \\    & & & \\   Continuous party measure &  & 0.475$^{***}$ & 0.009 \\    &  & (0.139) & (0.136) \\    & & & \\   Lag change in possessions &  & 0.267$^{***}$ & 0.285$^{***}$ \\    &  & (0.016) & (0.016) \\    & & & \\   Logged travel distance &  & $-$0.430$^{**}$ & 0.142 \\    &  & (0.191) & (0.187) \\    & & & \\   Number hours since last game &  & 0.012$^{*}$ & 0.017$^{**}$ \\    &  & (0.007) & (0.007) \\    & & & \\   Constant & $-$0.047 & 70.581$^{***}$ & 64.886$^{***}$ \\    & (0.090) & (2.267) & (2.220) \\    & & & \\  \hline \\[-1.8ex]  Observations & 15,658 & 6,426 & 6,426 \\  R$^{2}$ & 0.001 & 0.045 & 0.051 \\  Adjusted R$^{2}$ & 0.001 & 0.045 & 0.051 \\  Residual Std. Error & 11.159 (df = 15656) & 11.937 (df = 6421) & 11.688 (df = 6421) \\  F Statistic & 16.357$^{***}$ (df = 1; 15656) & 75.841$^{***}$ (df = 4; 6421) & 86.788$^{***}$ (df = 4; 6421) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

>>>>

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the metropolitan statistical area (MSA) 
in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the total number
of musical groups, sound recording studios, and musical publishers in county of the 
last game location, else emit zero.  We also re-scale this feature to lie within the
interval $[0,1]$ such that the magnitude of the coefficient is comparable with our
discrete indicator of nightlife.
This captures how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

\paragraph{How an unknown spread variable may (not) affect our inference}
For example, if bookkeepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house bookkeeper systematically mis-values
teams which are more likely to be encountered after playing a city with active nightlife,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reasonably reliable. 
So, if we do find an effect in our variable of interest,
we can be reasonably confident in our conclusion that there is a causal effect and not some other systematic bias.

\subsection{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the MSA-quarter level. We obtain this data for each metropolitan statistical area corresponding
to an NBA or MLB team from years 2010-2016. 
As a proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular county-year: we simply calculate the total number of establishments across all three music categories listed above. We caveat that since the Toronto Raptors and Toronto Blue Jays are located in Canada, 
we don't have BLS data for these teams.\footnote{We don't expect Toronto, Canada to be a party city.}

@@

\subsection{Results}
\subsubsection{NBA: using points-spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, recovery time, and home-team effect that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.
In laymens terms, teams under perform bookmaker's expectations after playing in cities 
with active nightlife.

% m0, m1
\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Meet the Spread} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]   Party discrete & $-$0.513$^{***}$ &  \\    & (0.145) &  \\    & & \\   Party continuous &  & $-$0.181$^{**}$ \\    &  & (0.091) \\    & & \\   Lag changes in posession & 0.0001 & 0.00004 \\    & (0.002) & (0.002) \\    & & \\   Logged travel distance & $-$0.017 & $-$0.014 \\    & (0.026) & (0.026) \\    & & \\   East-west travel direction & $-$0.145 & $-$0.113 \\    & (0.232) & (0.231) \\    & & \\   Number hours rest time & 0.001 & 0.001 \\    & (0.001) & (0.001) \\    & & \\   Time of game during day & 0.005 & 0.004 \\    & (0.013) & (0.013) \\    & & \\   Home team effect & 0.001 & 0.004 \\    & (0.043) & (0.043) \\    & & \\   Logged travel distance * east-west & 0.022 & 0.018 \\    & (0.036) & (0.036) \\    & & \\   Constant & $-$0.048 & $-$0.028 \\    & (0.378) & (0.378) \\    & & \\  \hline \\[-1.8ex]  Observations & 9,823 & 9,823 \\  Log Likelihood & $-$6,800.677 & $-$6,805.100 \\  Akaike Inf. Crit. & 13,619.350 & 13,628.200 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

\paragraph{Differences between NBA and MLB as it pertains to our analysis}
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win as both a function of the bookmakers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. Interviews with athletes suggest that MLB players are more likely to
restrict their partying to weekends.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up is repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

% m3, m4
\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Probability of Winning} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]   Continuous measure of nightlife & $-$0.126$^{*}$ &  \\    & (0.074) &  \\    & & \\   Nightlife (no weekend interaction) &  & $-$0.016 \\    &  & (0.110) \\    & & \\   Bookmaker's odds & 2.638$^{***}$ & 2.583$^{***}$ \\    & (0.163) & (0.165) \\    & & \\   Home-team effect & 0.008 & 0.034 \\    & (0.035) & (0.083) \\    & & \\   Number of rest days & 0.019 & 0.018 \\    & (0.020) & (0.020) \\    & & \\   Logged travel distance & $-$0.002 & $-$0.003 \\    & (0.005) & (0.005) \\    & & \\   Weekend & 0.044 & 0.003 \\    & (0.038) & (0.030) \\    & & \\   Constant & $-$1.344$^{***}$ & $-$1.324$^{***}$ \\    & (0.081) & (0.113) \\    & & \\  \hline \\[-1.8ex]  Observations & 24,734 & 24,033 \\  Log Likelihood & $-$16,946.080 & $-$16,470.850 \\  Akaike Inf. Crit. & 33,906.160 & 32,955.700 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

Indeed, we see that if players have enough recovery time after visiting cities such as Los
Angeles or New York, we cannot reject the null hypothesis that player performance is unaffected
by visitation to cities with active nightlife. The analagous placebo test
for MLB is not relevant, since almost all games are played back to back.

% mp
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   party placebo & 0.072 \\    & (0.082) \\    & \\   lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   east-west travel direction & $-$0.099 \\    & (0.232) \\    & \\   number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   time of game during day & 0.001 \\    & (0.013) \\    & \\   home team effect & $-$0.009 \\    & (0.043) \\    & \\   logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.070 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.702 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  
\subsection{Possible confounders or potential issues}

@@

% m2
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Demeanead Rebounds Per Minute \\  \hline \\[-1.8ex]   Continuous party measure & $-$0.003$^{*}$ \\    & (0.002) \\    & \\   Lag change in possessions & 0.00003 \\    & (0.0001) \\    & \\   Logged travel distance & 0.001 \\    & (0.001) \\    & \\   Number hours since last game & 0.00003 \\    & (0.00002) \\    & \\   Constant & $-$0.008$^{*}$ \\    & (0.004) \\    & \\  \hline \\[-1.8ex]  Observations & 43,547 \\  R$^{2}$ & 0.0003 \\  Adjusted R$^{2}$ & 0.0002 \\  Residual Std. Error & 0.104 (df = 43542) \\  F Statistic & 2.945$^{**}$ (df = 4; 43542) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

% mpa1, tpa, tpo
\hspace{-100pt}
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Points Admitted by Team & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.828$^{***}$ &  &  \\    & (0.699) &  &  \\    & & & \\   Continuous party measure &  & 1.667$^{***}$ & 0.080 \\    &  & (0.587) & (0.575) \\    & & & \\   Lag change in possessions &  & 0.267$^{***}$ & 0.285$^{***}$ \\    &  & (0.016) & (0.016) \\    & & & \\   Logged travel distance &  & $-$0.447$^{**}$ & 0.143 \\    &  & (0.191) & (0.187) \\    & & & \\   Number hours since last game &  & 0.011 & 0.017$^{**}$ \\    &  & (0.007) & (0.007) \\    & & & \\   Constant & $-$0.047 & 70.710$^{***}$ & 64.859$^{***}$ \\    & (0.090) & (2.271) & (2.223) \\    & & & \\  \hline \\[-1.8ex]  Observations & 15,658 & 6,426 & 6,426 \\  R$^{2}$ & 0.001 & 0.045 & 0.051 \\  Adjusted R$^{2}$ & 0.001 & 0.044 & 0.051 \\  Residual Std. Error & 11.159 (df = 15656) & 11.941 (df = 6421) & 11.688 (df = 6421) \\  F Statistic & 16.357$^{***}$ (df = 1; 15656) & 74.877$^{***}$ (df = 4; 6421) & 86.792$^{***}$ (df = 4; 6421) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}
